GVN-Hoist: Hoisting Computations from Branches

### 2016-08-05 hoist.tex
restructure sections

\section{Code hoisting}
Hoisting also reduces critical path length of execution in out of
order machines (but not in sequential machines), by exposing ILP
before the conditional where the instruction was hoisted.

@@

\subsection{Hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze
them for aliasing memory references. As long as all the operands are
available (which in SSA form we do, or may be there are chained
dependencies), the scalar can be hoisted. Care must be taken in case
of hoisting scalars too far, as that may increase register pressure
and result in spills. For example hoisting a scalar past a call. In
that case the call may result in save and restore of the register the
scalar may be defined to. In our current implementation we do not
hoist scalars past a call. Another way to mitigate this problem is be
to reinstantiate (rematerialize) the computation after a call (may be
as a different pass).

\subsection{Using MemorySSA}
\subsection{Iterative update to memory ssa during the hoisting process}

\subsection{Without MemorySSA using the dominance information}

\subsection{Hoisting loads}
In order to value number stores, we value number (hash) the address from where
the value is to be loaded. Then we need to compute the hoisting point just like
we did for the scalars. At the hoisting point we compute if the operands to the
load (an address) is available. If it is not available we can try to compute the
addresss if that is possible.  Along the path from current position of the load
instruction through the hoiting point, we find if there are aliasing stores, we
find if any of the basic blocks are landing pads or targets of indirect
branches. In those cases we discard the candidate. Since the load increases the
register pressure by one (unless the operand to load becomes dead at the load),
we evaluate if it is beneficial to hoist the load.

\subsection{Hoisting stores}
In order to value number stores, we value number (hash) the address as well as
the value to be stored to that address. For the legality, we need to check the
dependency requirements just like hoisting loads. We need to check that the
operands of store are available at the hoisting point, that there are no
aliasing loads or store along the path from the current position through the
hoisting point and, none of the basic blocks along the path are landing pads.
Once all the legality checks are satisfied, it is a good candidate to be
hoisted.

\subsection{Hoisting calls}
Calls can be hoisted by value-numbering all the arguments. For safety
we have to check that there are no other side-effects between a call and
its hoisting point, just like loads and stores.

\subsection{Reduce register pressure}
\label{hoist:reg-pressure}
Following example explains how code hoisting can actually reduce the register pressure.
Consider the following example where the labels prefixed with 'p' represent the position of
instruction in a basic block (names prefixed with bb).

\subsection{In the presence of calls}
Hoisting scalars across calls is tricky because it can increase the number of
spills. During the frame lowering of calls, the argument registers, in general,
the caller saved registers are saved because they might be modified by the
callee and after the call they are restored \cite{frame-lowering}. So before the
call, the register pressure is high because the number of available registers
are reduced by the number of caller saved registers. In that situation if a
computation is hoisted across the call, that would increase the total number of
registers required by one, thus contributing to the register pressure.

\subsection{Hoisting too far away}
When we load a value from memory, we require an extra register to keep the value
around. Hoisting a load means we need to keep the value alive for a longer
period. That could result in spills later in the register allocation.

\subsection{Comparative analysis of different approaches to hoisting}
\subsection{Triangle approach}
This is the most basic approach to code-hoisting when the instructions are
hoisted to immediate parent in a diamond like structure. When there are
identical computations in both the branches of a conditional, and all the
legality checks are met, the instruction can be hoisted to the predecessor basic
block. This approach is simple to implement and has minimal impact on the
register pressure.

\subsection{Pessimistic approach}
To extend the triangle approach to the entire function, the algorithm traverses
the basic blocks in the invers depth-first order. All the leaf nodes are visited
before the non-leaf nodes because instructions are hoisted upwards. We start
from a branch and collect the GVN of all the instructions. Then in the sibling
basic block we find if there are any identical computations. Once we find an
identical computation we try to hoist it to the nearest common dominator which
is the immediate predecessor of both the basic blocks. We traverse the basic
block from top to down so that the once an instruction is hoisted it creates
opportunities for other dependent instructions. After all the sibling basic
blocks have all the identical computations hoisted, we continue the hoisting
process with their parent and its sibling of their parent until we reach the
beginning of the function.

\subsection{Optimistic approach}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We see a lot of computations hoisted (see
Section~\ref{sec:experimental-results}. In order to do that, we collect the GVN
of all the instructions in the function. After that we iterate on the list of
instructions having identical GVN. After that we find the common dominator
dominating all the identical computations and try to hoist the instruction in
that basic block, if feasible. Many times it is not possible to hoist all the
instructions to one common dominator, due to legality/profitability
constraints. In those cases, the algorithmn would partion the list of
computations a partially hoist a set of computations to their common
dominator. For more details see Section~\ref{subsec:partition}.

\subsection{Hoisting bottom up with optimistic approach}
Optimistic means collect the set of all potential redundancies and
then discard the bad ones.

@@

\subsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
The set of candidates are sorted in the order of their DFS-in
numbers. So that candidates close to each other in the list are also
closer (??) in the CFG.

@@

>>>>

\newpage

\section{Code hoisting}

\subsection{Overview}

Hoisting also reduces critical path length of execution in out of
order machines (but not in sequential machines), by exposing ILP
before the conditional where the instruction was hoisted.

@@

\subsubsection{Hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze
them for aliasing memory references. As long as all the operands are
available (which in SSA form we do, or may be there are chained
dependencies), the scalar can be hoisted. Care must be taken in case
of hoisting scalars too far, as that may increase register pressure
and result in spills. For example hoisting a scalar past a call. In
that case the call may result in save and restore of the register the
scalar may be defined to. In our current implementation we do not
hoist scalars past a call. Another way to mitigate this problem is be
to reinstantiate (rematerialize) the computation after a call (may be
as a different pass).

\subsubsection{Hoisting loads}
In order to value number stores, we value number (hash) the address from where
the value is to be loaded. Then we need to compute the hoisting point just like
we did for the scalars. At the hoisting point we compute if the operands to the
load (an address) is available. If it is not available we can try to compute the
addresss if that is possible.  Along the path from current position of the load
instruction through the hoiting point, we find if there are aliasing stores, we
find if any of the basic blocks are landing pads or targets of indirect
branches. In those cases we discard the candidate. Since the load increases the
register pressure by one (unless the operand to load becomes dead at the load),
we evaluate if it is beneficial to hoist the load.

\subsubsection{Hoisting stores}
In order to value number stores, we value number (hash) the address as well as
the value to be stored to that address. For the legality, we need to check the
dependency requirements just like hoisting loads. We need to check that the
operands of store are available at the hoisting point, that there are no
aliasing loads or store along the path from the current position through the
hoisting point and, none of the basic blocks along the path are landing pads.
Once all the legality checks are satisfied, it is a good candidate to be
hoisted.

\subsubsection{Hoisting calls}
Calls can be hoisted by value-numbering all the arguments. For safety
we have to check that there are no other side-effects between a call and
its hoisting point, just like loads and stores.

\subsection{Algorithm}
A good full page.

\subsubsection{Hoisting bottom up with optimistic approach}
Optimistic means collect the set of all potential redundancies and
then discard the bad ones.

@@

\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
The set of candidates are sorted in the order of their DFS-in
numbers. So that candidates close to each other in the list are also
closer (??) in the CFG.

@@

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Following example explains how code hoisting can actually reduce the register pressure.
Consider the following example where the labels prefixed with 'p' represent the position of
instruction in a basic block (names prefixed with bb).

\subsubsection{In the presence of calls}
Hoisting scalars across calls is tricky because it can increase the number of
spills. During the frame lowering of calls, the argument registers, in general,
the caller saved registers are saved because they might be modified by the
callee and after the call they are restored \cite{frame-lowering}. So before the
call, the register pressure is high because the number of available registers
are reduced by the number of caller saved registers. In that situation if a
computation is hoisted across the call, that would increase the total number of
registers required by one, thus contributing to the register pressure.

\subsubsection{Hoisting too far away}
When we load a value from memory, we require an extra register to keep the value
around. Hoisting a load means we need to keep the value alive for a longer
period. That could result in spills later in the register allocation.

\subsection{Other Approaches}

\subsubsection{Triangle approach}
This is the most basic approach to code-hoisting when the instructions are
hoisted to immediate parent in a diamond like structure. When there are
identical computations in both the branches of a conditional, and all the
legality checks are met, the instruction can be hoisted to the predecessor basic
block. This approach is simple to implement and has minimal impact on the
register pressure.

\subsubsection{Pessimistic approach}
To extend the triangle approach to the entire function, the algorithm traverses
the basic blocks in the invers depth-first order. All the leaf nodes are visited
before the non-leaf nodes because instructions are hoisted upwards. We start
from a branch and collect the GVN of all the instructions. Then in the sibling
basic block we find if there are any identical computations. Once we find an
identical computation we try to hoist it to the nearest common dominator which
is the immediate predecessor of both the basic blocks. We traverse the basic
block from top to down so that the once an instruction is hoisted it creates
opportunities for other dependent instructions. After all the sibling basic
blocks have all the identical computations hoisted, we continue the hoisting
process with their parent and its sibling of their parent until we reach the
beginning of the function.

\subsubsection{Optimistic approach}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We see a lot of computations hoisted (see
Section~\ref{sec:experimental-results}. In order to do that, we collect the GVN
of all the instructions in the function. After that we iterate on the list of
instructions having identical GVN. After that we find the common dominator
dominating all the identical computations and try to hoist the instruction in
that basic block, if feasible. Many times it is not possible to hoist all the
instructions to one common dominator, due to legality/profitability
constraints. In those cases, the algorithmn would partion the list of
computations a partially hoist a set of computations to their common
dominator. For more details see Section~\ref{subsec:partition}.

\subsubsection{PRE based hoisting}

\newpage

\subsection{Comparative analysis of different approaches to hoisting}
